Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-10 pecomplexity.tex
some editing of the results section

In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% refereed to as the ``learning/training" signal, $\{X_{i,obs}\}_{i=1}^{i=n}$ and the last 10\% known as the ``test/correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to build the initial models (e.g., LMA,ARIMA,na\"ive) as described in Section \ref{sec:compModel}. The test signal is used both to test the forecasts and to refit the models after forecasting. In particular, using the models we perform $k$ 1-step predictions $\{p_j\}_{j=n+1}^{k+n+1}$, after each prediction we append the training signal with the the next point in the correct signal. We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.

@@

$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}$$

@@

\begin{table}[htdp]
\caption{Average MASE for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average wpe at word length 5 and 6 for each signal. }
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
\hline
\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\

@@

As can be seen in Table \ref{tab:error} the relationship between prediction accuracy and the weighted permutation entropy (WPE) is much as we conjectured: performance traces with high WPE are
indeed harder to predict using the forecasting models described in Section \ref{sec:compModel}. Figure~\ref{fig:pred_vs_ent} demonstrates this primary finding, with
the exception of one outlier whose behavior we can explain. In Figure~\ref{fig:pred_vs_ent} we plot the best prediction (i.e., the lowest MASE over methods on that signal) for each of the 8 signals.  
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$. The
  dashed line is a least-squares linear fit of all the points except for {\tt
  SVD$_1$} which we have excluded for reasons explained in the text.}
  \label{fig:pred_vs_ent}
\end{figure}
We find that the relationship between the two is roughly linear. The single outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE: namely
that in the absence of any large features, as all the other regimes have, the
WPE effectively falls back to the standard PE for which the noisy behavior
drives toward 1.0.

This primary finding: ``Signals that are more complex are harder to predict." At first glance this finding may seem incredibly trivial, of course a complex signal is harder to predict than a trivial one, but this is \emph{not} the core result here as we will elucidate now with Figure \ref{fig:lma_vs_arima}.
\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
    \caption{The MASE of LMA vs weighted permutation entropy. For each of these,
    the word length used is $6$.}
    \label{fig:lma_pred_vs_ent}
  \end{subfigure}%
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \caption{The MASE of ARIMA vs weighted permutation entropy. For each of
    these, the word length used is $6$.}
    \label{fig:arima_pred_vs_ent}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
    \caption{The MASE values for LMA against ARIMA. The dashed line is the
    identity, delineating the traces for which either LMA or ARIMA performed
    better. All traces except those from {\tt SVD$_1$} lie above the line,
    indicating that LMA is better suited prediction method for the computer
   traces considered.}\label{fig:lma_vs_arima}

  %\end{subfigure}
\end{figure}

\item Talk about structure analysis. Just because there is forward infromation transfer, does not mean that linear predictors can get at this. 
\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by side plot. 
\item  full results.  Image: MASE vs. WPE for both LMA \&
ARIMA.  Points to make: 
\begin{enumerate}
\item clusters are distributed differently
\item clusters are shaped differently---tight or not 
\item clusters move
differently between LMA and ARIMA 
\item finally, the diagonal line is
important.  If you're below it, you could do better.

@@

In Figures~\ref{fig:lma_pred_vs_ent} and \ref{fig:arima_pred_vs_ent} we directly
compare the performance of the LMA and ARIMA prediction methods (respectively)
to the value of the weighted permutation entropy for all runs of each program
under consideration. The LMA MASE values are largely similar to those of the
best predictions, primarily because LMA often performed superior to ARIMA and
the naive method. On the other hand, the ARIMA MASE values are largely
uncorrelated with WPE values. The fact that ARIMA is uncorrelated with WPE is in large part one of the major findings of this work. More specifically, say we tried to predict an arbitrary noisy real-valued time series with an ``out-of-the-box" prediction strategy like ARIMA as is proposed as a good strategy in \cite{ARIMA-autofit} and say you got inconsistent and bad forecasts, (i.e., perform worse than the na\"ive random walk strategy ($MASE>1$). How do we know if the prediction strategy is not adequte for the prediction task or if the signal is simply too complex to predict. If a signal is too complex and little forward information transfer is present we may not be able to do better than the random walk, in which case we should not kill ourselves over finding a more complicated prediction strategy. However, if we measure the complexity to be low, $WPE<0.85$ we can most likely do much better than the random walk and should search for more adequate prediction strategies. For example, consider \col (the blue +s in Figure \ref{fig:lma_vs_arima}). If we use the out-of-the-box ARIMA from \cite{ARIMA-autofit} we get MASE ranging from 0.5 all the way to 1.0. That is we may think that the best we can do is twice as good as random-walk but sometimes we only do as well as random walk. In which case the random walk prediction strategy is probably your best bet as it is so simple and produces very similar error. However, if we calculate WPE we see that \col has a WPE of $0.5131\pm 0.0034$. This incredibly low entropy implies a great deal of deterministic structure that we can utilize for prediction, even though this did not appear to be the case with ARIMA. In fact using LMA on this signal we get an average MASE of $0.05\pm0.0018$, an error that is on-average 20 times better than the random walk forecast. Alternatively, we can look at the other end of the spectrum at \gcc. This signal has a WPE of $0.9430\pm0.0013$ and each prediction method we applied was significantly out performed by a simple random walk (the best strategy in the case of structurless time series. Since \gcc has such little structure according to WPE it is safe to assume that he best strategy will be random walk and that contining to search for a more accurate prediction strategy will most likely be a fruitless process.

@@

The structure being picked up on by WPE may be linear or nonlinear. When you have a low WPE and a high ARIMA it could be that the structure WPE is picking up is simply nonlinear structure tha-1t LMA can handle but ARIMA cannot. So while the ARIMA prediction look realy bad there is plenty of structure present as suggested by WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it into account and does bad.

>>>>

In order to analyze correctness of each prediction we split each time series
into two pieces: the first 90\% refereed to as the ``learning/training" signal,
$\{X_{i,obs}\}_{i=1}^{i=n}$ and the last 10\% known as the ``test/correct"
signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to build the
initial models (e.g., LMA, ARIMA, na\"ive) as described in Section
\ref{sec:compModel}. The test signal is used both to test the forecasts and to
refit the models after forecasting. In particular, using the models we perform
$k$ 1-step predictions $\{p_j\}_{j=n+1}^{k+n+1}$, after each prediction we
append the training signal with the the next point in the correct signal. We
would like to note that this rebuilding occurs due to a problem with ARIMA
models converging to a mean prediction if too long of a prediction horizon is
used, this is not a handicap of either LMA or na\"ive.

@@

\begin{align*}
  \textrm{MASE} = \sum_{j=n+1}^{k+n+1}\frac{|c_j-p_j| }{\frac{k}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|}
\end{align*}

@@

\begin{table}[htdp]
\caption{Average MASE for 1-step predictions at a 10\% prediction horizon over 15 runs for each signal and average WPE at word length 5 and 6 for each signal. }
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
                   & MASE LMA    & MASE ARIMA &MASE na\"{i}ve & $l=5$  & $l=6$ \\
\hline
\gcc                  & $ 1.5296\pm 0.0214$ & $1.8366 \pm0.0157 $ & $1.7970\pm0.0095$& $0.9510 \pm 0.0011$ & $0.9430 \pm 0.0013$ \\

@@

As can be seen in Table \ref{tab:error} the relationship between prediction
accuracy and the weighted permutation entropy (WPE) is much as we conjectured:
performance traces with high WPE are indeed harder to predict using the
forecasting models described in Section \ref{sec:compModel}.
Figure~\ref{fig:pred_vs_wpe} demonstrates this primary finding, with the
exception of one outlier whose behavior we can explain. In
Figure~\ref{fig:pred_vs_wpe} we plot the best prediction (i.e., the lowest MASE
over methods on that signal) for each of the 8 signals.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/prediction_vs_entropy}
  \caption{The best MASE among all runs and prediction methods vs weighted
  permutation entropy. For each of these, the word length used is $6$. The
  dashed line is a least-squares linear fit of all the points except for {\tt
  SVD$_1$} which we have excluded for reasons explained in the text.}
  \label{fig:pred_vs_wpe}
\end{figure}
We find that the relationship between the two is roughly linear. The single
outlier, \svd regime 1, does not fit the trend due to a weakness in the WPE:
namely that in the absence of any large features, as all the other regimes have,
the WPE effectively falls back to the standard PE which the noisy behavior
drives toward 1.0.

This primary finding: ``Signals that are more complex are harder to predict." At
first glance this finding may seem incredibly trivial, of course a complex
signal is harder to predict than a trivial one, but this is \emph{not} the core
result here as we will elucidate now with Figure \ref{fig:predictions_vs_wpe}.
\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/LMA_prediction_vs_entropy}
    \caption{The MASE of LMA vs weighted permutation entropy. For each of these,
    the word length used is $6$.}
    \label{fig:lma_pred_vs_wpe}
  \end{subfigure}%
  \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \caption{The MASE of ARIMA vs weighted permutation entropy. For each of
    these, the word length used is $6$.}
    \label{fig:arima_pred_vs_wpe}
  \end{subfigure}
  % \begin{subfigure}{0.5\textwidth}
  %   \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
  %   \caption{The MASE values for LMA against ARIMA. The dashed line is the
  %   identity, delineating the traces for which either LMA or ARIMA performed
  %   better. All traces except those from {\tt SVD$_1$} lie above the line,
  %   indicating that LMA is better suited prediction method for the computer
  %  traces considered.}\label{fig:lma_vs_arima}
  % \end{subfigure}
  \label{fig:predictions_vs_wpe}
\end{figure}

\item Talk about structure analysis. Just because there is forward information
transfer, does not mean that linear predictors can get at this.
\item For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by
side plot.
\item full results.  Image: MASE vs. WPE for both LMA \& ARIMA.  Points to make:
\begin{enumerate}
\item clusters are distributed differently
\item clusters are shaped differently---tight or not
\item clusters move differently between LMA and ARIMA
\item finally, the diagonal line is important. If you're below it, you could do
better.

@@

In Figures~\ref{fig:lma_pred_vs_wpe} and \ref{fig:arima_pred_vs_wpe} we directly
compare the performance of the LMA and ARIMA prediction methods (respectively)
to the value of the weighted permutation entropy for all runs of each program
under consideration. The LMA MASE values are largely similar to those of the
best predictions, primarily because LMA often performed superior to ARIMA and
the naive method. On the other hand, the ARIMA MASE values are largely
uncorrelated with WPE values. The fact that ARIMA is uncorrelated with WPE is in
large part one of the major findings of this work. More specifically, say we
tried to predict an arbitrary noisy real-valued time series with an ``out-of-
the-box" prediction strategy like ARIMA as proposed in \cite{ARIMA-autofit}, and
say you got inconsistent and bad forecasts, (i.e., perform worse than the
na\"ive random walk strategy (MASE$>1$). How do we determine if the prediction
strategy is not adequate for the prediction task, or if the signal is simply too
complex to predict. If a signal is too complex and little forward information
transfer is present we may simply not be possible to do better than the random
walk, in which case we should not worry ourselves over finding a more
complicated prediction strategy. However if we measure the complexity to be low,
$\textrm{WPE}<0.85$ (see Fig.~\ref{fig:pred_vs_wpe}), we can likely do better
than the random walk and should search for more adequate prediction strategies.

By way of example, consider \col (the blue +s in
Figure~\ref{fig:predictions_vs_wpe}). If we use the out-of-the-box ARIMA method
from \cite{ARIMA-autofit} we get MASE values ranging from $\sim0.5$ to
$\sim1.0$. From that information we may assume that the best predictions that
can be done are twice as good as random-walk, and in some cases predictions will
only do as well as random-walk. In the latter such cases, the random-walk
prediction strategy is probably the best bet as it is so simple and produces
very similar error. However, if we calculate the WPE we see that \col has a WPE
of $0.5131 \pm 0.0034$. This incredibly low entropy implies a great deal of
deterministic structure which can be utilize for prediction, even though this
did not appear to be the case with ARIMA. In fact using LMA on this signal we
get an average MASE value of $0.05\pm0.0018$, an error that is on average 20
times better than the random walk forecast, and 10 times better than the best
ARIMA forcast, with little variance. Alternatively, we can look at the other end
of the spectrum: \gcc. This signal has a WPE of $0.9430\pm0.0013$ and each
prediction method we applied was significantly out performed by a simple random
walk (the best strategy in the case of structureless time series). Since \gcc
has such little structure according to WPE it is safe to assume that he best
strategy will be random walk and that continuing to search for a more accurate
prediction strategy will likely be a fruitless process.

@@

The WPE is sensitive to both linear and nonlinear structure. When you have a low
WPE and a high ARIMA it could be that the structure WPE is picking up is simply
nonlinear structure that LMA can handle but ARIMA cannot. So while the ARIMA
prediction look really bad there is plenty of structure present as suggested by
WPE and taken advantage of by LMA but since it is nonlinear ARIMA can't take it
into account and does bad.
